---
title: "Stats Modeling and Computing Final Project Intermediate Report"
author: "Avi Kodali(ank93) and Vatsal Malkari(vm593)"
date: "2026-04-01"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(boot)
```

Our dataset has approximately 2,200 customers and has information about their demographics (age, gender, etc), purchasing behavior (how often they purchase, what products, etc), and marketing campaign responses (success rate of each campaign marketing, etc). Each observation corresponds to a single customer, with features that capture their personal characteristics and interaction with the company.

Feature Categories 
Demographics: Age (Year_Birth can be converted to Age), Income, Education level, Marital status, Number of children. This is important as it can affect purchasing power and preferences.

Spending Behavior: MntWines (amount spent on wine), MntMeatProducts (amount spent on meat), MntFruits (amount spent on fruits), MntFishProducts (amount spent on fish), MntSweetProducts (amount spent on sweets), MntGoldProds (amount spent on gold products. These will help us with K-means clustering.

Activity: Recency (days since last purchase), NumWebPurchases (web purchases), NumCatalogPurchases (catalog purchases), NumStorePurchases (store purchases), NumWebVisitsMonth (web visits per month). These show how active each customer is. We can also group customers by each of these columns to see which purchase is more effective for which demographic.

Marketing Response: AcceptedCmp1 to AcceptedCmp5 (shows 0 if customer did not respond to campaign and 1 if they did yes), Response (what the data last campaign showed). We will use it to see how successful campaigns have been. This helps with bayesian inference (probability of response) and causal inference (treatment effect)


```{r}
df <- read.csv("marketing_campaign.csv", sep = "\t", stringsAsFactors = FALSE)
```

# Data Preprocessing

```{r}
df_clean <- df %>%
  mutate(
    Age = 2014 - Year_Birth,
    Income = ifelse(is.na(Income), median(Income, na.rm = TRUE), Income)
  ) %>%
  filter(Age >= 18 & Age <= 100) %>%
  select(-Z_CostContact, -Z_Revenue) 
df_clean %>% head()
```


```{r}
spending_cols <- c("MntWines", "MntFruits", "MntMeatProducts", "MntFishProducts", "MntSweetProducts", "MntGoldProds")
pca_result <- prcomp(df_clean[, spending_cols], center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
df_clean$Spend_PC1 <- pca_result$x[, 1]
df_clean$Spend_PC2 <- pca_result$x[, 2]
```

```{r}
numeric_features <- c("Age", "Spend_PC1", "Spend_PC2",
                      "Recency", "NumWebPurchases", "NumCatalogPurchases",
                      "NumStorePurchases", "NumWebVisitsMonth")

non_pc <- setdiff(numeric_features, c("Spend_PC1", "Spend_PC2", "Spend_PC3"))
df_scaled <- df_clean %>%
  mutate(across(all_of(non_pc), ~ as.numeric(scale(.)))) %>% 
  select(-Income)
```


```{r}
set.seed(123)
wss <- sapply(1:12, function(k) {
  kmeans(df_scaled[, numeric_features],
         centers = k, nstart = 25, iter.max = 50)$tot.withinss
})

plot(1:12, wss, type="b", xlab="Number of Clusters (k)", ylab="Within cluster Sum of Squares",
     main=" K-means", pch=19, col="blue")
text(1:12, wss, labels=round(wss,1), pos=3, cex=0.8)
wss_diff <- diff(wss)
wss_diff
```

```{r}
set.seed(123)
kmeans_result <- kmeans(df_scaled[, numeric_features], centers = 4, nstart = 25, iter.max = 50)

df_clean$Cluster <- kmeans_result$cluster

df_clean %>%
  group_by(Cluster) %>%
  summarize(
    Avg_Age = mean(Age),
    Avg_Income = mean(Income),
    Avg_Spend_PC1 = mean(Spend_PC1),
    Avg_Spend_PC2 = mean(Spend_PC2),
    Avg_Recency = mean(Recency),
    Avg_WebPurchases = mean(NumWebPurchases),
    Avg_CatalogPurchases = mean(NumCatalogPurchases),
    Avg_StorePurchases = mean(NumStorePurchases),
    Avg_WebVisits = mean(NumWebVisitsMonth),
    Avg_Response = mean(Response)
  )
```

```{r}
# Average income by cluster
df_clean %>%
  group_by(Cluster) %>%
  summarize(Avg_Income = mean(Income)) %>%
  ggplot(aes(x = factor(Cluster), y = Avg_Income, fill = factor(Cluster))) +
  geom_col() +
  labs(title = "Average Income by Cluster", x = "Cluster", y = "Income") +
  theme_minimal()
```


```{r}
ggplot(df_clean, aes(x = Spend_PC1, y = Spend_PC2, color = factor(Cluster))) +
geom_point(alpha = 0.6) + labs(title = "Customer Clusters", x = "Spend_PC1", y = "Spend_PC2", color = "Cluster")
```

```{r}
cluster_summary <- df_clean %>% group_by(Cluster) %>%
summarize( Responded = sum(Response), Total = n())

cluster_summary
```

```{r}
high_value_clusters <- c(1, 2, 4)

resp_rate <- function(data, indices) {
  mean(data[indices])
}

# Bootstrap for each cluster
bootstrap_results <- list()

for(i in 1:4) {
  cluster_data <- df_clean %>% filter(Cluster == i) %>% pull(Response)
  boot_obj <- boot(cluster_data, resp_rate, R = 5000)
  boot_ci <- boot.ci(boot_obj, type = "perc")
  
  bootstrap_results[[i]] <- data.frame(Cluster = i, Response_Rate = boot_obj$t0,
    Boot_SE = sd(boot_obj$t), CI_Lower = boot_ci$percent[4], CI_Upper = boot_ci$percent[5])
}

bootstrap_summary <- do.call(rbind, bootstrap_results)
print(bootstrap_summary)
```


```{r}
cluster_summary <- df_clean %>% group_by(Cluster) %>% summarize( Responded = sum(Response), 
    Total = n(), Response_Rate = Responded / Total )
cluster_summary
```



```{r}
response_rate <- function(data, indices) mean(data[indices])

set.seed(123)
per_cluster_bootstrap <- list()

for (i in 1:4) {
  cluster_responses <- df_clean %>% filter(Cluster == i) %>% pull(Response)
  boot_obj <- boot(cluster_responses, response_rate, R = 5000)
  ci_perc <- boot.ci(boot_obj, type = "perc")$percent[4:5]

  per_cluster_bootstrap[[i]] <- data.frame(
    Cluster       = i,
    N_Customers   = length(cluster_responses),
    Response_Rate = boot_obj$t0,
    Boot_SE       = sd(boot_obj$t),
    CI_Lower      = ci_perc[1],
    CI_Upper      = ci_perc[2]
  )
}

cluster_response_summary <- do.call(rbind, per_cluster_bootstrap)
print(cluster_response_summary)

# Visualize per-cluster CIs
ggplot(cluster_response_summary, aes(x = factor(Cluster), y = Response_Rate)) +
  geom_point(size = 4, colour = "steelblue") +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper),
                width = 0.2, colour = "steelblue") +
  geom_hline(yintercept = mean(df_clean$Response),
             linetype = "dashed", colour = "red") +
  labs(
    title    = "Per-Cluster Response Rates with 95% Bootstrap CIs",
    subtitle = "Red dashed = overall response rate",
    x        = "Cluster",
    y        = "Response Rate"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
high_value_responses  <- df_clean$Response[df_clean$Cluster %in% high_value_clusters]
other_cluster_responses <- df_clean$Response[!(df_clean$Cluster %in% high_value_clusters)]

n_high_value  <- length(high_value_responses)
n_other       <- length(other_cluster_responses)

observed_rate_diff <- mean(high_value_responses) - mean(other_cluster_responses)

set.seed(123)
B <- 5000
bootstrap_rate_diffs <- numeric(B)

for (i in 1:B) {
  high_value_resample  <- sample(high_value_responses,    n_high_value, replace = TRUE)
  other_resample       <- sample(other_cluster_responses, n_other,      replace = TRUE)
  bootstrap_rate_diffs[i] <- mean(high_value_resample) - mean(other_resample)
}

# All three CI methods (lecture slides 19, 20, 21)
alpha            <- 0.05
bootstrap_se     <- sd(bootstrap_rate_diffs)

ci_normal <- c(
  observed_rate_diff - qnorm(1 - alpha/2) * bootstrap_se,
  observed_rate_diff + qnorm(1 - alpha/2) * bootstrap_se
)

ci_percentile <- quantile(bootstrap_rate_diffs, c(alpha/2, 1 - alpha/2))

ci_pivotal <- c(
  2 * observed_rate_diff - quantile(bootstrap_rate_diffs, 1 - alpha/2),
  2 * observed_rate_diff - quantile(bootstrap_rate_diffs, alpha/2)
)

cat("\n=== TWO-SAMPLE BOOTSTRAP: HIGH-VALUE vs OTHER CLUSTERS ===\n")
cat("Mean response rate (high-value clusters):", round(mean(high_value_responses), 4), "\n")
cat("Mean response rate (other clusters):     ", round(mean(other_cluster_responses), 4), "\n")
cat("Observed difference in rates:            ", round(observed_rate_diff, 4), "\n")
cat("Bootstrap SE:                            ", round(bootstrap_se, 4), "\n\n")

cat("Normal     95% CI: [", round(ci_normal[1], 4),     ",", round(ci_normal[2], 4),     "]\n")
cat("Percentile 95% CI: [", round(ci_percentile[1], 4), ",", round(ci_percentile[2], 4), "]\n")
cat("Pivotal    95% CI: [", round(ci_pivotal[1], 4),    ",", round(ci_pivotal[2], 4),    "]\n\n")

# Decision rule (lecture slide 50): does the CI exclude zero?
significant <- ci_percentile[1] > 0
cat("Decision rule: percentile CI excludes 0 →", significant, "\n")
if (significant) {
  cat("=> Statistically significant. High-value targeting outperforms.\n")
}

# Visualize the bootstrap distribution of the difference
ggplot(data.frame(rate_diff = bootstrap_rate_diffs), aes(x = rate_diff)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = observed_rate_diff, colour = "red", linewidth = 1) +
  geom_vline(xintercept = ci_percentile, colour = "darkgreen",
             linetype = "dashed", linewidth = 0.8) +
  geom_vline(xintercept = 0, colour = "black", linetype = "dotted") +
  labs(
    title    = "Bootstrap Distribution: Difference in Response Rates",
    subtitle = paste0("Red = observed (", round(observed_rate_diff, 3),
                      ") | Green dashed = 95% percentile CI | Dotted = null"),
    x        = "Difference in Response Rate (high-value, other)",
    y        = "Count"
  ) +
  theme_minimal(base_size = 13)
```


```{r}
n_contacted_targeted   <- sum(df_clean$Cluster %in% high_value_clusters)
n_contacted_mass       <- nrow(df_clean)

set.seed(123)
B <- 5000
targeted_strategy_rate <- numeric(B)
mass_strategy_rate     <- numeric(B)

for (i in 1:B) {
  high_value_resample <- sample(high_value_responses,
                                n_high_value, replace = TRUE)
  full_dataset_resample <- sample(df_clean$Response,
                                  n_contacted_mass, replace = TRUE)
  targeted_strategy_rate[i] <- mean(high_value_resample)
  mass_strategy_rate[i]     <- mean(full_dataset_resample)
}

cat("\n=== STRATEGY EFFICIENCY (responses per customer contacted) ===\n")
cat("Mass strategy (target all):    ", round(mean(mass_strategy_rate), 4),
    " 95% CI: [", round(quantile(mass_strategy_rate, 0.025), 4),
    ",", round(quantile(mass_strategy_rate, 0.975), 4), "]\n")
cat("Targeted (high-value clusters):", round(mean(targeted_strategy_rate), 4),
    " 95% CI: [", round(quantile(targeted_strategy_rate, 0.025), 4),
    ",", round(quantile(targeted_strategy_rate, 0.975), 4), "]\n\n")

# Headline numbers for the report
expected_responses_mass     <- mean(mass_strategy_rate)     * n_contacted_mass
expected_responses_targeted <- mean(targeted_strategy_rate) * n_contacted_targeted

cat("Customers contacted — mass strategy:    ", n_contacted_mass, "\n")
cat("Customers contacted — targeted strategy:", n_contacted_targeted,
    " (", round(100 * n_contacted_targeted / n_contacted_mass, 1),
    "% of total)\n")
cat("Expected responses  — mass strategy:    ", round(expected_responses_mass), "\n")
cat("Expected responses  — targeted strategy:", round(expected_responses_targeted),
    " (", round(100 * expected_responses_targeted / expected_responses_mass, 1),
    "% of mass-targeting)\n")
```

```{r}
df_clean <- df_clean %>%
  mutate(
    Targeted = as.integer(Cluster %in% high_value_clusters),
    Income_Group = case_when(
      Income < 40000  ~ "Low",
      Income <= 70000 ~ "Medium",
      TRUE            ~ "High"
    )
  )
```

```{r}
income_vs_targeted <- df_clean %>%
  group_by(Income_Group) %>%
  summarise(Targeting_Rate = round(mean(Targeted), 3), N = n())

income_vs_response <- df_clean %>%
  group_by(Income_Group) %>%
  summarise(Response_Rate = round(mean(Response), 3), N = n())

cat("Income → Targeting rate:\n");  print(income_vs_targeted)
cat("\nIncome → Response rate:\n"); print(income_vs_response)
cat("\nIncome predicts both. Stratification is required.\n")

# Raw (biased) difference in means
raw_means <- df_clean %>%
  group_by(Targeted) %>%
  summarise(Mean_Response = mean(Response), N = n())

raw_diff <- diff(raw_means$Mean_Response)
cat("\nRaw difference in means (biased):", round(raw_diff, 4), "\n")
```

```{r}
strata <- c("Low", "Medium", "High")

stratum_results <- lapply(strata, function(g) {
  d  <- df_clean %>% filter(Income_Group == g)
  y1 <- d$Response[d$Targeted == 1]
  y0 <- d$Response[d$Targeted == 0]
  n1 <- length(y1)
  n0 <- length(y0)

  psi_hat <- mean(y1) - mean(y0)
  se_g    <- sqrt(var(y1)/n1 + var(y0)/n0)   # within-stratum Neyman SE
  ci_lo   <- psi_hat - 1.96 * se_g
  ci_hi   <- psi_hat + 1.96 * se_g

  cat("\nIncome group:", g, "\n")
  cat("  n_treated =", n1, " | n_control =", n0, "\n")
  cat("  Mean(Y | A=1) =", round(mean(y1), 4),
      " | Mean(Y | A=0) =", round(mean(y0), 4), "\n")
  cat("  ATE estimate:", round(psi_hat, 4),
      " | 95% CI: [", round(ci_lo, 4), ",", round(ci_hi, 4), "]\n")

  data.frame(
    Income_Group = g,
    n_treated = n1, n_control = n0,
    mean_y1 = mean(y1), mean_y0 = mean(y0),
    ATE = psi_hat,
    SE  = se_g,
    CI_lower = ci_lo, CI_upper = ci_hi
  )
})

stratum_df <- bind_rows(stratum_results) %>%
  mutate(
    N_stratum = n_treated + n_control,
    Weight    = N_stratum / sum(N_stratum)
  )

# Weighted overall ATE — uses within-stratum SEs (correct formula)
overall_ATE   <- sum(stratum_df$ATE * stratum_df$Weight)
overall_SE    <- sqrt(sum(stratum_df$Weight^2 * stratum_df$SE^2))
overall_CI_lo <- overall_ATE - 1.96 * overall_SE
overall_CI_hi <- overall_ATE + 1.96 * overall_SE

cat("\n=== OVERALL STRATIFIED ATE ===\n")
cat("Weighted ATE: ", round(overall_ATE, 4), "\n")
cat("95% CI:       [", round(overall_CI_lo, 4),
    ",", round(overall_CI_hi, 4), "]\n")
cat("Interpretation: Conditional on income, being in a high-value cluster\n")
cat("increases response probability by approximately",
    round(overall_ATE * 100, 1), "percentage points.\n")

# Restricted estimand: drop High-income stratum due to positivity violation
stratum_df_overlap <- stratum_df %>% filter(Income_Group != "High")

# Re-weight the remaining strata so they sum to 1
weights_restricted <- stratum_df_overlap$N_stratum / sum(stratum_df_overlap$N_stratum)

overall_ATE_restricted <- sum(stratum_df_overlap$ATE * weights_restricted)
overall_SE_restricted  <- sqrt(sum(weights_restricted^2 * stratum_df_overlap$SE^2))
overall_CI_lo_r <- overall_ATE_restricted - 1.96 * overall_SE_restricted
overall_CI_hi_r <- overall_ATE_restricted + 1.96 * overall_SE_restricted

cat("\n=== RESTRICTED STRATIFIED ATE (Low + Medium only) ===\n")
cat("Weighted ATE:    ", round(overall_ATE_restricted, 4), "\n")
cat("95% CI: [", round(overall_CI_lo_r, 4), ",", round(overall_CI_hi_r, 4), "]\n")
cat("Estimand: ATE among customers with Income < $70K\n")
cat("(", sum(stratum_df_overlap$N_stratum), "of", sum(stratum_df$N_stratum), "customers)\n")
```

```{r}
# Restrict to Low + Medium for both data and strata
df_overlap <- df_clean %>% filter(Income_Group != "High")
strata_overlap <- c("Low", "Medium")

# Modified version of compute_weighted_ate that uses strata_overlap
compute_weighted_ate_overlap <- function(data) {
  s <- lapply(strata_overlap, function(g) {
    d <- data %>% filter(Income_Group == g)
    y1 <- d$Response[d$Targeted == 1]
    y0 <- d$Response[d$Targeted == 0]
    if (length(y1) == 0 || length(y0) == 0) {
      return(data.frame(ATE = 0, N = nrow(d)))
    }
    data.frame(ATE = mean(y1) - mean(y0), N = nrow(d))
  })
  s <- bind_rows(s)
  w <- s$N / sum(s$N)
  sum(s$ATE * w)
}

t_obs_overlap <- abs(compute_weighted_ate_overlap(df_overlap))

set.seed(123)
B <- 5000
t_perm_overlap <- numeric(B)

for (b in 1:B) {
  df_perm <- df_overlap %>%
    group_by(Income_Group) %>%
    mutate(Targeted = sample(Targeted)) %>%
    ungroup()
  t_perm_overlap[b] <- abs(compute_weighted_ate_overlap(df_perm))
}

p_value_overlap <- mean(t_perm_overlap >= t_obs_overlap)
cat("Restricted permutation p-value (Low + Medium only):", 
    round(p_value_overlap, 4), "\n")
```

```{r}
# Stratum-level ATE plot
ggplot(stratum_df, aes(x = Income_Group, y = ATE)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_errorbar(aes(ymin = CI_lower, ymax = CI_upper),
                width = 0.15, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  geom_hline(yintercept = overall_ATE, linetype = "dotted",
             colour = "darkgreen", linewidth = 0.8) +
  labs(
    title    = "Stratum-Level ATE Estimates with 95% CIs",
    subtitle = paste0("Red dashed = null (ATE=0)  |  ",
                      "Green dotted = overall weighted ATE (",
                      round(overall_ATE, 3), ")"),
    x = "Income Group",
    y = "Estimated ATE (probability scale)"
  ) +
  theme_minimal(base_size = 13)

# Permutation distribution — make sure observed statistic is on-chart
hist(t_perm_overlap,
     breaks = 60,
     xlim   = c(0, max(t_obs_overlap, max(t_perm_overlap)) * 1.05),
     col    = "lightblue",
     border = "white",
     main   = "Permutation Distribution Under Sharp Null",
     xlab   = "|Weighted ATE|",
     sub    = paste0("Red dashed = observed (", round(t_obs_overlap, 4),
                     ")  |  p-value = ", round(p_value_overlap, 4)))
abline(v = t_obs_overlap, col = "red", lwd = 2, lty = 2)
```

```{r}
cat("\n=== FINAL SUMMARY ===\n")
cat("Bootstrap evidence:\n")
cat("  - High-value clusters (1, 2, 3) response rate:",
    round(mean(high_value_responses) * 100, 1), "%\n")
cat("  - Other clusters (2, 4) response rate:    ",
    round(mean(other_cluster_responses) * 100, 1), "%\n")
cat("  - Difference in response rates:           ",
    round(observed_rate_diff, 4), "\n")
cat("  - Percentile 95% CI for difference:       [",
    round(ci_percentile[1], 4), ",", round(ci_percentile[2], 4), "]\n")
cat("\nCausal evidence:\n")
cat("  - Stratified ATE (controlling for income):",
    round(overall_ATE, 4), "\n")
cat("  - 95% CI for stratified ATE:              [",
    round(overall_CI_lo, 4), ",", round(overall_CI_hi, 4), "]\n")
cat("  - Permutation p-value:                    ",
    round(p_value_overlap, 4), "\n")
```


```{r}
# lin estimator
df_clean <- df_clean %>%
  mutate(Edu_Num = as.integer(factor(Education,
                                levels = c("Basic", "2n Cycle", "Graduation", "Master", "PhD"))),
    Married = as.integer(Marital_Status %in% c("Married", "Together")),
    Children = Kidhome + Teenhome
  )

# confounder set
covars <- c("Income", "Age", "Edu_Num", "Married", "Children", "Recency",
            "NumWebPurchases", "NumCatalogPurchases", "NumStorePurchases", "NumWebVisitsMonth")

# Demean covariates (Lin 2013)
df_model <- df_clean %>%
  select(Response, Targeted, all_of(covars)) %>%
  drop_na() %>%
  mutate(across(all_of(covars), ~ . - mean(.), .names = "c_{.col}"))

centered <- paste0("c_", covars)

# interacted regression (Lin estimator)
formula_lin <- as.formula(
  paste("Response ~ Targeted * (", paste(centered, collapse = " + "), ")")
)

lin_fit <- lm(formula_lin, data = df_model)

# Robust SE
library(sandwich)
library(lmtest)
robust <- coeftest(lin_fit, vcov = vcovHC(lin_fit, type = "HC2"))

ate     <- coef(lin_fit)["Targeted"]
se      <- robust["Targeted", "Std. Error"]
ci_low  <- ate - 1.96 * se
ci_high <- ate + 1.96 * se

cat("\n  regression-adjusted ATE (Lin estimator)  \n")
cat("ATE:    ", round(ate, 4), "\n")
cat("SE:     ", round(se, 4), "\n")
cat("95% CI: [", round(ci_low, 4), ",", round(ci_high, 4), "]\n")
cat("p-value:", round(robust["Targeted", "Pr(>|t|)"], 4), "\n")
```

```{r}
ggplot(ate_comparison %>% filter(!is.na(CI_Lower)),
       aes(x = Method, y = ATE)) +
  geom_point(size = 3, colour = "steelblue") +
  geom_errorbar(aes(ymin = CI_Lower, ymax = CI_Upper), width = 0.2, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "red") +
  coord_flip() + labs(title = "ATE Estimates Across Adjustment Methods",
    subtitle = "Effect shrinks after adjusting for recency & activity",
    x = NULL, y = "Estimated ATE") + theme_minimal(base_size = 13)
```

After adjusting for key demographic and behavioral variables using the Lin estimator, the estimated treatment effect of 
being in a high‑value cluster drops to 2.97%, with a 95% confidence interval from –4.23% to 10.16% and a p‑value of 0.4188. 
This means that once we control for factors like age, income, education, marital status, number of children, recency, and purchase‑channel activity, 
the remaining difference in response rates between targeted and non‑targeted customers is small and statistically indistinguishable from zero. 
In plain language, the cluster label itself does not appear to cause higher response rates after accounting for these underlying behavioral differences.
This result clarifies why the high‑value clusters performed better: they contain customers who were already more engaged, more recent, 
and more active across purchase channels. The clusters are still extremely valuable for targeting because they efficiently group together 
the customers most likely to respond, even if the cluster label itself is not the causal driver. Tthe regression‑adjusted analysis shows 
that selective advertising still works. It works because the clusters capture high‑engagement behavior, not because the cluster assignment 
independently increases response probability.
