---
title: "Stats Modeling and Computing Final Project Intermediate Report"
author: "Avi Kodali(ank93) and Vatsal Malkari(vm593)"
date: "2026-04-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

Our dataset has approximately 2,200 customers and has information about their demographics (age, gender, etc), purchasing behavior (how often they purchase, what products, etc), and marketing campaign responses (success rate of each campaign marketing, etc). Each observation corresponds to a single customer, with features that capture their personal characteristics and interaction with the company.

Feature Categories 
Demographics: Age (Year_Birth can be converted to Age), Income, Education level, Marital status, Number of children. This is important as it can affect purchasing power and preferences.

Spending Behavior: MntWines (amount spent on wine), MntMeatProducts (amount spent on meat), MntFruits (amount spent on fruits), MntFishProducts (amount spent on fish), MntSweetProducts (amount spent on sweets), MntGoldProds (amount spent on gold products. These will help us with K-means clustering.

Activity: Recency (days since last purchase), NumWebPurchases (web purchases), NumCatalogPurchases (catalog purchases), NumStorePurchases (store purchases), NumWebVisitsMonth (web visits per month). These show how active each customer is. We can also group customers by each of these columns to see which purchase is more effective for which demographic.

Marketing Response: AcceptedCmp1 to AcceptedCmp5 (shows 0 if customer did not respond to campaign and 1 if they did yes), Response (what the data last campaign showed). We will use it to see how successful campaigns have been. This helps with bayesian inference (probability of response) and causal inference (treatment effect)


```{r}
df <- read.csv("marketing_campaign.csv", sep = "\t", stringsAsFactors = FALSE)
```

# Data Preprocessing

```{r}
df_clean <- df %>%
  mutate(
    Age = 2014 - Year_Birth,
    Income = ifelse(is.na(Income), median(Income, na.rm = TRUE), Income)
  )%>%   filter(Age >= 18 & Age <= 100)
df_clean
```


```{r}
spending_cols <- c("MntWines", "MntFruits", "MntMeatProducts", "MntFishProducts", "MntSweetProducts", "MntGoldProds")
pca_result <- prcomp(df_clean[, spending_cols], center = TRUE, scale. = TRUE)
summary(pca_result)
```

```{r}
df_clean$Spend_PC1 <- pca_result$x[, 1]
df_clean$Spend_PC2 <- pca_result$x[, 2]
df_clean$Spend_PC3 <- pca_result$x[, 3]
```


```{r}
df_kmeans <- df_clean %>%
  mutate(across(where(is.numeric), ~ as.numeric(scale(.))))
df_kmeans
```


```{r}
numeric_features <- c("Age", "Income", "Spend_PC1", "Spend_PC2", "Recency", "NumWebPurchases", "NumCatalogPurchases", "NumStorePurchases", "NumWebVisitsMonth")

df_scaled <- df_clean %>%
  mutate(across(all_of(numeric_features), ~ as.numeric(scale(.))))
```


```{r}
set.seed(123)
wss <- sapply(1:12, function(k) {
  kmeans(df_scaled[, numeric_features], centers = k, nstart = 25)$tot.withinss
})

plot(1:12, wss, type="b", xlab="Number of Clusters (k)", ylab="Within cluster Sum of Squares",
     main=" K-means", pch=19, col="blue")
text(1:12, wss, labels=round(wss,1), pos=3, cex=0.8)
wss_diff <- diff(wss)
wss_diff
```

```{r}
set.seed(123)
kmeans_result <- kmeans(df_scaled[, numeric_features], centers = 4, nstart = 25)

df_clean$Cluster <- kmeans_result$cluster

df_clean %>%
  group_by(Cluster) %>%
  summarize(
    Avg_Age = mean(Age),
    Avg_Income = mean(Income),
    Avg_Spend_PC1 = mean(Spend_PC1),
    Avg_Spend_PC2 = mean(Spend_PC2),
    Avg_Recency = mean(Recency),
    Avg_WebPurchases = mean(NumWebPurchases),
    Avg_CatalogPurchases = mean(NumCatalogPurchases),
    Avg_StorePurchases = mean(NumStorePurchases),
    Avg_WebVisits = mean(NumWebVisitsMonth),
    Avg_Response = mean(Response)
  )
```

```{r}
# Average income by cluster
df_clean %>%
  group_by(Cluster) %>%
  summarize(Avg_Income = mean(Income)) %>%
  ggplot(aes(x = factor(Cluster), y = Avg_Income, fill = factor(Cluster))) +
  geom_col() +
  labs(title = "Average Income by Cluster", x = "Cluster", y = "Income") +
  theme_minimal()
```


```{r}
ggplot(df_clean, aes(x = Spend_PC1, y = Spend_PC2, color = factor(Cluster))) +
geom_point(alpha = 0.6) + labs(title = "Customer Clusters", x = "Spend_PC1", y = "Spend_PC2", color = "Cluster")
```

```{r}

library(dplyr)
cluster_summary <- df_clean %>% group_by(Cluster) %>%
summarize( Responded = sum(Response), Total = n())

cluster_summary
```

```{r}
# we apply bootstrap here
library(boot)

resp_rate <- function(data, indices) {
  mean(data[indices])
}

# Bootstrap for each cluster
bootstrap_results <- list()

for(i in 1:4) {
  cluster_data <- df_clean %>% filter(Cluster == i) %>% pull(Response)
  boot_obj <- boot(cluster_data, resp_rate, R = 5000)
  boot_ci <- boot.ci(boot_obj, type = "perc")
  
  bootstrap_results[[i]] <- data.frame(Cluster = i, Response_Rate = boot_obj$t0,
    Boot_SE = sd(boot_obj$t), CI_Lower = boot_ci$percent[4], CI_Upper = boot_ci$percent[5])
}

bootstrap_summary <- do.call(rbind, bootstrap_results)
print(bootstrap_summary)
```


```{r}
cluster_summary <- df_clean %>% group_by(Cluster) %>% summarize( Responded = sum(Response), 
    Total = n(), Response_Rate = Responded / Total )
cluster_summary
```



```{r}
# bootstrap simulation
B <- 5000
targeted_clusters <- c(1, 2, 3)

# Run simulation
set.seed(123)

results_all <- numeric(B)
results_targeted <- numeric(B)

for(i in 1:B) {
  # all customers
  sim_all <- df_clean %>%
    sample_n(size = nrow(df_clean), replace = TRUE) %>%
    summarise(responses = sum(Response))
  results_all[i] <- sim_all$responses
  
  # only selected clusters
  sim_targeted <- df_clean %>%
    filter(Cluster %in% targeted_clusters) %>%
    sample_n(size = nrow(.), replace = TRUE) %>%
    summarise(responses = sum(Response))
  results_targeted[i] <- sim_targeted$responses
}

#  results
mean(results_all)
mean(results_targeted)
sd(results_all)
sd(results_targeted)

# we see the differences between clusters here
df_clean %>%
  group_by(Cluster) %>%
  summarize(
    Avg_Income = mean(Income),
    Avg_Recency = mean(Recency),
    Avg_WebVisits = mean(NumWebVisitsMonth),
    Avg_StorePurchases = mean(NumStorePurchases),
    Avg_CatalogPurchases = mean(NumCatalogPurchases)
  )
```


```{r}
# Density plot of simulation results
sim_results <- data.frame( Responses = c(results_all, results_targeted),
  Strategy = rep(c("Target All", "Target Selected"), each = B))

ggplot(sim_results, aes(x = Responses, fill = Strategy)) +
  geom_density(alpha = 0.5) +
  labs(title = "Bootstrap Simulation: Campaign Outcomes", x = "Number of Responses", y = "Density") 
```

```{r}
# test if targeting causes higher response rates, or if it's just because targeted customers have higher incomes.
low_threshold <- quantile(df_clean$Income, 1/3)
high_threshold <- quantile(df_clean$Income, 2/3)

df_clean$Income_Group <- "Medium"
df_clean$Income_Group[df_clean$Income <= low_threshold] <- "Low"
df_clean$Income_Group[df_clean$Income >= high_threshold] <- "High"

# targeted customers
df_clean$Targeted <- "No"
df_clean$Targeted[df_clean$Cluster %in% c(1, 2, 3)] <- "Yes"

# response rates
df_clean %>%
  group_by(Income_Group, Targeted) %>%
  summarize(Response_Rate = mean(Response), 
            Customers = n())
```

```{r}
# causal evidence that targeting works, even when you compare customers with the same income
causal_table <- df_clean %>%
  group_by(Income_Group, Targeted) %>%
  summarise(Response_Rate = mean(Response), 
            Customers = n())

print(causal_table)

ggplot(causal_table, aes(x = Income_Group, y = Response_Rate, fill = Targeted)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Response Rates by Income Group and Targeting Status",
       x = "Income Group", y = "Response Rate", fill = "Targeted") 
```


```{r}
# confidence intervals for the difference in response rates between targeted and non-targeted customers, within each income group
# low income group
yes_low <- df_clean$Response[df_clean$Income_Group == "Low" & df_clean$Targeted == "Yes"]
no_low <- df_clean$Response[df_clean$Income_Group == "Low" & df_clean$Targeted == "No"]
prop.test(c(sum(yes_low), sum(no_low)), c(length(yes_low), length(no_low)))$conf.int

# medium income group
yes_med <- df_clean$Response[df_clean$Income_Group == "Medium" & df_clean$Targeted == "Yes"]
no_med <- df_clean$Response[df_clean$Income_Group == "Medium" & df_clean$Targeted == "No"]
prop.test(c(sum(yes_med), sum(no_med)), c(length(yes_med), length(no_med)))$conf.int

# high income group
yes_high <- df_clean$Response[df_clean$Income_Group == "High" & df_clean$Targeted == "Yes"]
no_high <- df_clean$Response[df_clean$Income_Group == "High" & df_clean$Targeted == "No"]
prop.test(c(sum(yes_high), sum(no_high)), c(length(yes_high), length(no_high)))$conf.int
```
